\PassOptionsToPackage{euenc}{fontspec}
\documentclass[
			en,
			phd,
			draft,
			garamond,
			12pt,
			]{sty/swathesis}
	

% Intro			 
% Howto			30 (17)
% IDS				40 (27)
% SQLODB		30 (12)
% StackDB		20 (10)
% rel				 5
% Conc			 5


%\newcommand{\tmp}[1]{#1}
\newcommand{\tmpStart}{\color{DarkBlue}}
\newcommand{\tmpEnd}{\color{Black}}
%\newcommand{\tmpStart}{}
%\newcommand{\tmpEnd}{}
%\includeonly{introduction}
%\includeonly{introduction, interactive_slicing}
%\includeonly{odb_sql}
%\includeonly{relwork_conclusion}
%\includeonly{cross_system}

%\usepackage{parskip}
\input{setup}
%\newcommand{\defaultbaselinestretch}{1.4}
%\renewcommand{\baselinestretch}{\defaultbaselinestretch}
\doublespacing

\iftrue
\addbibresource{references.bib}
\addbibresource{publications.bib}
\else
\bibliography{references}
\bibliography{publications}
\fi
\DeclareBibliographyCategory{mine}
\addtocategory{mine}{treffer12:understanding_code_with_execution,treffer13:extending_omniscient_debuggers,pape14:extending_a_java_virtual,treffer14:dynamic_slicing_with_soot,treffer14:omniscient_debugging_in_database,treffer15:time-travel_queries_for_omniscient,treffer16:the_slice_navigator_focused,treffer17:back-in-time_debugging_in_heterogeneous,treffer17:bringing_back-in-time_debugging,treffer18:interactive_dynamic_slicing_tracking}
\nocite{treffer12:understanding_code_with_execution,treffer13:extending_omniscient_debuggers,pape14:extending_a_java_virtualtreffer14:dynamic_slicing_with_soot,treffer14:omniscient_debugging_in_database,treffer15:time-travel_queries_for_omniscient,treffer16:the_slice_navigator_focused,treffer17:back-in-time_debugging_in_heterogeneous,treffer17:bringing_back-in-time_debugging,treffer18:interactive_dynamic_slicing_tracking}

\setcounter{tocdepth}{1}
\setcounter{secnumdepth}{3}

\renewcommand*{\glsnamefont}[1]{\textbf{#1}}


\usepackage{titlepage}
\TitlePageStyle[discipline=phd, whom={der}, subject=degree]{hpi-swa}
%\degree{Doktor der Naturwissenschaften\\ (Doctor rerum naturalium)}
\degree{Doktor der Ingenieurwissenschaften\\ (Doktor-Ingenieur)}
\faculty{%
Digital Engineering Fakultät\\
der Universität Potsdam}

\advisor{Prof.\,Dr.\,Hasso Plattner}


% ABGABEDATUM
% \setdate{2012}{04}{05}
% \date{\datedate}

\author{Arian Treffer}
\location{Berlin}
\extratitle{\raggedleft Arian Treffer, \\ Omniscient Debugging in Enterprise Applications\par}
\title{Omniscient Debugging in Enterprise Applications}
%\subtitle{\smaller{Tracking Erroneous State in Heterogeneous Software Systems}}
\othertitle{Omniscient Debugging in Unternehmensanwendungen}
%\othersubtitle{Ein Untertitel ist Optional}


%% remove in final version
%\usepackage[colorinlistoftodos]{todonotes}
%\renewcommand{\todo}[2][\empty]{\pdfmargincomment[color=orange,icon=Note,subject=TODO,#1]{#2}}
%\usepackage{pdfcomment}
%\newcommand{\todo}[2][]{\pdfmargincomment[color=orange,icon=Note,subject={TODO},author={#1}]{#2}}


\newacronym{afl}{AFL}{automatic fault-localization}
\newacronym{pdg}{PDG}{program dependence graph}
\newacronym{ddg}{DDG}{dynamic dependence graph}
\newacronym{odb}{ODB}{omniscient debugger}
\newacronym{orm}{ORM}{object-relational mapper}
\newacronym{erp}{ERP}{enterprise resource planning}
%\DeclareAcronym{afl}{short=AFL,long=automated fault-localization}
%\DeclareAcronym{pdg}{short=PDG,long=program dependence graph}
%\DeclareAcronym{ddg}{short=DDG,long=dynamic dependence graph}
%\acrodef{afl}{automated fault-localization}
%\acrodef{pdg}{program dependence graph}
%\acrodef{ddg}{dynamic dependence graph}

\nocite{*}
%\makeglossaries
\begin{document}

\frontmatter
\begin{otherlanguage}{german}
\maketitle
\end{otherlanguage}

\linenumbers

\pdfbookmark[1]{Abstract}{toc:abstract}
\begin{abstract}
%Developers spend up to 60 percent of their time working on maintenance tasks.
%Much of that time is spend in the debugger, searching for the cause of an observed problem.
%Modern debugging tools can increase developer productivity by reducing the time needed to locate faults.
%Omniscient debuggers, for instance, allow developers to follow the infection chain backwards in time, which removes the overhead of frequently having to restart debug sessions.
%Alas, omniscient debugging, like many other advanced debugging techniques, is not widely used in practice.

%In this thesis, we present

In the development of enterprise applications, debugging is notoriously difficult.
Many enterprise applications are large and complex software systems built with a layered architecture, where each layer may be developed in different programming languages, run in different environments, and therefore require different tools to be debugged.
They work with large amounts of data, which can also be a source of errors, but is stored in a database outside the debugger's scope.
%
%
Modern debugging tools can increase developer productivity by reducing the time needed to locate faults.
Omniscient debuggers, for instance, allow developers to follow the infection chain backwards in time, which removes the overhead of frequently having to restart debug sessions.
%Alas, omniscient debugging, like many other advanced debugging techniques, is not widely used in practice.
However, while omniscient debugging can improve debugging in general,
it does not address the specific problems of developers debugging enterprise applications.

%Modern debugging techniques do not meet the needs of developers of such systems.

To effectively debug enterprise applications, developers need a debugger that supports tracking infection chains across sub-system boundaries, helps to navigate through large amounts of code, and includes the database as part of the application state.
We developed methods to adapt omniscient debugging to meet these needs.

%In our work, we take necessary steps towards a debugger that allows developers to efficiently debug long infection chains in a heterogeneous system.

\emph{Interactive Dynamic Slicing} is a new debugging workflow combining back-in-time debugging and dynamic slicing.
A new configurable dynamic slicing algorithm allows developers to iteratively reconfigure slicing criteria to minimize the search space of active code without interrupting the debugging session.
The \emph{Slice Navigator}, a complementary debugging tool, enables developers to use this workflow effectively.
Combining debugger and slicer in one view, the Slice Navigator summarizes relevant program state, serves as an alternative to breakpoints, and allows developers to quickly adjust the slice.
In a user study, we found that developers using Interactive Dynamic Slicing needed 40\% less time to locate faults than with a regular debugger and 15\% less time than with an omniscient debugger.


%We combine omniscient debugging with dynamic analysis, in particular dynamic slicing, 
%to reduce the time needed to track erroneous state over long distances.
 %processes the debugger's runtime data.
%The results are shown in the Slice Navigator, a novel UI component enriching the debugging
%experience in multiple ways. First, it supports developers’ short term memory
%by providing a summary of relevant program state and context for the current
%instruction. Second, it provides an alternative to breakpoints as it can be used
%to control the debugger to jump to related instructions, such as the last change of
%a variable. Third, it allows to directly reconfigure the slicing criteria, enabling
%developers to minimize the search space of active code without interrupting the
%debugging workflow.


%consisting of multiple components, such as client code, application layer, and database scripts.
%These components are 

%One area that could benefit greatly from better debugging tools is the development of enterprise applications.
%

%Multi-tier enterprise applications often 
%When debugging such systems, developers can be faced with long infection chains, the paths of erroneous states leading from the fault in the code to the observed failure, that may even cross sub-system boundaries.
%Following an infection chain can be a tedious and time-consuming task,
%,
 %but when execution crosses a system boundary developers have to switch tools and manually map state, such as request arguments, from one system to the next.
%This creates an additional mental burden and distracts from the actual problem.

%
%Enterprise applications use a relational database to handle complex operations on large amounts of data.
%While most modern databases allow to attach a debugger to queries or scripts, the general tool support for understanding complex programs in databases is rather poor compared to that of popular object-oriented languages.

To accommodate the demand for better debugging tools in database programs, we brought omniscient debugging down to the database.
We present a method to leverage the mix of imperative control flow and declarative SQL statements in stored procedures. 
This allows us to create an omniscient debugger much more efficient with large amounts of data than previous solutions for object-oriented languages.
Using \emph{Back-in-time SQL}, developers can submit queries against previous states of the database,
compare data from multiple points in time, and search directly for changes in the data.

Finally, we extended our database debugger to include the full application stack,
allowing developers the seamless debugging of the entire stack of a 3-tier application, from the front-end running in the browser down to stored procedures in the database. 
High-level visualizations guide developers in understanding sub-system interactions and help to sequentially debug concurrent requests.
%the relations between sub-systems and help to sequentially debug concurrent requests. 
An evaluation with a development project at SAP showed the feasibility of our approach.
We received encouraging developer feedback indicating that our database-inclusive full-stack debugger significantly simplifies the debugging of multi-tier enterprise applications.

%we were able to address 
%currently missing in the debugging toolset for enterprise applications.

\end{abstract}

%\begin{acronym}[XXX]
%\acro{afl}[AFL]{automated fault-localization}
%
%\acro{ddg}[DDG]{dynamic dependence graph}
%\end{acronym}

\pdfbookmark[1]{Zusammenfassung}{toc:zusammenfassung}
\begin{otherlanguage}{german}
\begin{zusammenfassung}
  Abstract auf deutsch
\end{zusammenfassung}
\end{otherlanguage}

\defaultstatement

\cleardoublepage
\pdfbookmark[1]{Contents}{toc:contents}
\tableofcontents
%\printglossary[type=main]

\mainmatter
\lstMakeShortInline[basicstyle=\ttfamily,language={Inline}]´

\include{introduction}

\include{interactive_slicing}

\include{odb_sql}

\include{cross_system}

\include{relwork_conclusion}

%\printglossaries


\cleardoublepage
%\phantomsection
\pdfbookmark[0]{Lists}{toc:listof}
\makeatletter
\addtocontents{toc}{\protect\contentsline{chapter}{Lists}{}{\@currentHref}}
\makeatother
\addcontentsline{toc}{section}{\listfigurename}
\listoffigures

\cleardoublepage
\phantomsection
\addcontentsline{toc}{section}{\listtablename}
\listoftables

\cleardoublepage
\phantomsection
\addcontentsline{toc}{section}{List of Listings}%\lstlistingname
\lstlistoflistings

%\cleardoublepage
%\addcontentsline{toc}{section}{Acronyms}
%\listofacronyms


\emergencystretch=3em
\newrefcontext[sorting=ynt]
\printbibliography[heading=bibintoc,title={Publications},category=mine]
\newrefcontext[sorting=nyt]
\printbibliography[heading=bibintoc,category=cited,notcategory=mine]

\printbibliography[title={TODO},notcategory=cited,notcategory=mine]

\end{document}
